--- 
title: "A short introduction to circular data analysis"
author: "Jolien Cremers"
date: "February 8th, 2020"
institute: | 
  | Section of Biostatistics, University of Copenhagen
  | Statistics Denmark
  | jcre@sund.ku.dk
output:
  # ioslides_presentation:
  #   citation_package: natbib
  #   css: css/style.css
  #   transition: 0
  #   self_contained: TRUE
  beamer_presentation:
    keep_tex: true
    latex_engine: pdflatex
    
csl: apa.csl
pandoc_args: --natbib
bibliography: bibliography_circular.bib

header-includes:
   - \usepackage{colortbl}   
   - \usepackage{tikz}
   - \usepackage{amsmath}
   - \usetikzlibrary{positioning}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r echo = FALSE}
library(bpnreg)
library(circular)
library(tikzDevice)
library(plotrix)
library(dplyr)
library(haven)
library(bpnreg)
```

```{r, echo = FALSE}
data <- read.table("DatatoSamplerPointingNorth.txt", header = TRUE, sep = "\t", dec = ".")
data$sex <- factor(data$sex, labels = c("male", "female"))
```

# What is circular data?

A variable $\theta$ defined on $[0, 2\pi)$ radians or $[0, 360)$ degrees:

\begin{figure}[H]
        \centering

\includegraphics[width=\textwidth]{plots/Circular.png}

		\label{}
\end{figure}


# Three different types of circular data

* Directions in 2D space / Compass data

* Time of day / clock times

* Circular measurement scales


# Directions in 2D space / compass data
\small

Spatial cognition:

* Brunyé, T. T., Burte, H., Houck, L. A., & Taylor, H. A. (2015). The map in our head is not oriented north: Evidence from a real-world environment.

* Postma, A., Zuidhoek, S., Noordzij, M. L., & Kappers, A. M. (2008). Keep an eye on your hands: on the role of visual mechanisms in processing of haptic space.

Eye movements:

* Mulder, K., Klugkist, I., van Renswoude, D., & Visser, I. (2020). Mixtures of peaked power Batschelet distributions for circular data with application to saccade directions.

# Time of day / clock times
\small

Phase difference:

* Puglisi, G., Leonetti, A., Landau, A., Fornia, L., Cerri, G., & Borroni, P. (2017). The role of attention in human motor resonance.


Time of day:

* Rocchi, M. B., & Perlini, C. (2002). Is the time of suicide a random choice? A new statistical perspective. 

* Brunsdon, C., & Corcoran, J. (2006). Using circular statistics to analyse time patterns in crime incidence. 

Circadian rhythms:

* Peeters, F., Berkhof, J., Delespaul, P., Rottenberg, J., & Nicolson, N. A. (2006). Diurnal mood variation in major depressive disorder.



# Circular measurement scales
\small

Rose of Leary:

  * Leary, T. (1958). Interpersonal diagnosis of personality. 
  
Interpersonal circumplex:

  * Gurtman, M. B. (2009). Exploring personality with the interpersonal circumplex.

\begin{figure}[H]


\includegraphics[width=0.4\textwidth]{plots/Leary.png}

		\label{}
\end{figure}

# Why do we need different analysis methods?

\begin{figure}[H]
        \centering

\includegraphics[width=\textwidth]{plots/Circular.png}

		\label{}
\end{figure}


# Circular summary statistics

```{r meanrbar, echo = FALSE}
par(mfrow = c(2,2),
    oma=c(0,0,0,0),
    mar=c(0,0,0,0),
    las=1, mgp=c(4,1,0),
    xaxs="r", yaxs="r")
plot(0, 0,  asp = 1, bty='n', axes = FALSE, xlab = "", ylab = "",  xlim = c(-1, 3), ylim = c(-1, 2))
plotrix::draw.circle(0, 0, 1)
arrows(c(0, 0, 0), c(0, 0, 0), cos(c(1, 0.5, 5)), sin(c(1, 0.5, 5)), length = 0.05)
plot(0, 0,  asp = 1, bty='n', axes = FALSE, xlab = "", ylab = "",  xaxs="r", yaxs="r", xlim = c(-1, 3), ylim = c(-1, 2))
plotrix::draw.circle(0, 0, 1)
arrows(c(0, cos(1), cos(1) + cos(0.5)),
       c(0, sin(1), sin(1) + sin(0.5)),
       c(cos(1), cos(1) + cos(0.5), cos(1) + cos(0.5) + cos(5)),
       c(sin(1), sin(1) + sin(0.5), sin(1) + sin(0.5) + sin(5)),
       length = 0.05)
arrows(0, 0, cos(1) + cos(0.5) + cos(5), sin(1) + sin(0.5) + sin(5), lty = 2, length = 0.05)

plot(0, 0,  asp = 1, bty='n', axes = FALSE, xlab = "", ylab = "",  xaxs="r", yaxs="r", xlim = c(-1, 3), ylim = c(-1, 2))
plotrix::draw.circle(0, 0, 1)
arrows(c(0, 0, 0), c(0, 0, 0), cos(c(0.3, 0.4, 0.5)), sin(c(0.3, 0.4, 0.5)), length = 0.05)
plot(0, 0,  asp = 1, bty='n', axes = FALSE, xlab = "", ylab = "",  xaxs="r", yaxs="r", xlim = c(-1, 3), ylim = c(-1, 2))
plotrix::draw.circle(0, 0, 1)
arrows(c(0, cos(0.3), cos(0.3) + cos(0.4)),
       c(0, sin(0.3), sin(0.3) + sin(0.4)),
       c(cos(0.3), cos(0.3) + cos(0.4), cos(0.3) + cos(0.4) + cos(0.5)),
       c(sin(0.3), sin(0.3) + sin(0.4), sin(0.3) + sin(0.4) + sin(0.5)),
       length = 0.05)
arrows(0, 0, cos(0.3) + cos(0.4) + cos(0.5), sin(0.3) + sin(0.4) + sin(0.5), lty = 2, length = 0.05)
``` 

# Circular mean direction

$$\bar{\theta} = \begin{cases} \tan^{-1}\left(\frac{\sum_{i=1}^n\sin\theta_i}{\sum_{i=1}^n\cos\theta_i}\right) & \sum_{i=1}^n\sin\theta_i > 0, \sum_{i=1}^n\cos\theta_i > 0 \\\tan^{-1}\left(\frac{\sum_{i=1}^n\sin\theta_i}{\sum_{i=1}^n\cos\theta_i}\right) + \pi & \sum_{i=1}^n\cos\theta_i < 0 \\ \tan^{-1}\left(\frac{\sum_{i=1}^n\sin\theta_i}{\sum_{i=1}^n\cos\theta_i}\right) + 2\pi & \sum_{i=1}^n\sin\theta_i < 0, \sum_{i=1}^n\cos\theta_i > 0\end{cases}$$

# Mean resultant length

$$R = \sqrt{\left(\sum_{i=1}^n\cos\theta_i\right)^2 + \left(\sum_{i=1}^n\sin\theta_i\right)^2}$$
$$\bar{R} = R/n$$

1 = All datapoints have the same value

0 $\neq$ uniform, e.g. bimodality at exactly opposite parts of the circle (0$^\circ$, 180$^\circ$) also leads to a zero value.

# Other summary statistics

* circular variance, $V = 1-\bar{R}$
* circular standard deviation $v = {-2\log(1-V)}^{0.5}$
* circular median direction 
* circular modal direction

see @fisher1995statistical for an overview.

# Circular distributions

Three classes:

* Intrinsic distributions (e.g. von Mises)
* Wrapped distributions ($\theta \sim N(\mu, \sigma)\: \text{mod}\: 2\pi$)
* Projected distributions ($\theta = (\sin\theta, \cos\theta)^t = \boldsymbol{y} \sim N_2(\boldsymbol{\mu}, \boldsymbol{\Sigma})/||\boldsymbol{y}||$)

# Intrinsic distributions

Directly defined on the circle, e.g.:

* von Mises distribution

$$f(\theta) = [2\pi I_0(\kappa)]^{-1} \exp[\kappa\cos(\theta-\mu)]$$

mean direction = $\mu$

concentration parameter = $\kappa$

mean resultant length = $A_1(\kappa)= I_1(\kappa)/I_0(\kappa)$, see @fisher1995statistical


# Wrapped distributions

\begin{figure}
\centering
\includegraphics[width = \textwidth]{plots/wrapping.png}
\label{wrapping}
\end{figure}

# Projected distributions

\begin{figure}
\centering
\includegraphics[width = \textwidth]{plots/embedding4.pdf}
\label{embedding4}
\end{figure}


# Projected distributions

Instead of inference on the circular outcome, $\theta$, directly we perform 
inference on underlying bivariate data, $\boldsymbol{\text{Y}} \sim N(\boldsymbol{\mu}, \boldsymbol{I})$.

* Advantage: \textbf{flexible} approach, make use of already existing models for 
bivariate linear data.

* Disadvantage: $\boldsymbol{\text{Y}}$ and $r$ are latent, need special 
inference methods (EM algorithm [@presnell1998projected] or \textbf{data
augmented MCMC sampling} [@nunez2005bayesian]).

* Disadvantage: \textbf{interpretation} on two components instead of on
circle.


# A short example analysis using the R-package bpnreg

\small
\vspace{0.5cm}

Brunyé, T. T., Burte, H., Houck, L. A., & Taylor, H. A. (2015). The map in our head is not oriented north: Evidence from a real-world environment. PLoS One, 10(9), e0135803.

```{r, echo = FALSE}
par(mfrow = c(1,2))

plot(as.circular(data %>% filter(sex == "male") %>% select(theta)), stack = TRUE, main = "males")
plot(as.circular(data %>% filter(sex == "female") %>% select(theta)), stack = TRUE, main = "females")
```


# Fit a circular regression model
```{r, results = FALSE }

fit <- bpnr(theta ~ sex + age + experience + sbsod,
            data = data,
            its = 10000, burn = 100,
            seed = 101)

```

# Results

Estimated circular means for males and females with average age, experience and Santa Barbara Sense of Direction (SBSOD) score:

\vspace{0.5cm}

\scriptsize
```{r, results = FALSE}
coef_circ(fit, type = "categorical", units = "degrees")
```


```{r, echo = FALSE}
res <- coef_circ(fit, type = "categorical", units = "degrees")

# Means
round(res$Means, 2)
```

# Results

Estimated mean resultant length for males and females with average age, experience and Santa Barbara Sense of Direction (SBSOD) score:

\vspace{0.5cm}

\scriptsize
```{r, echo = FALSE}
a1 <- fit$B1[,"(Intercept)"]
a2 <- fit$B2[,"(Intercept)"]
b1 <- fit$B1[,"sexfemale"]
b2 <- fit$B2[,"sexfemale"]

zeta   <- sqrt((a1)^2 + (a2)^2)^2/4
spread <- sqrt((pi * zeta)/2) * exp(-zeta) * (besselI(zeta, 0) + besselI(zeta, 1))
zeta2   <- sqrt((a1 + b1)^2 + (a2 + b2)^2)^2/4
spread2 <- sqrt((pi * zeta2)/2) * exp(-zeta2) * (besselI(zeta2, 0) + besselI(zeta2, 1))

out <- rbind(c(mean(spread), mode_est(spread), sd(spread), hpd_est(spread)),
             c(mean(spread2), mode_est(spread2), sd(spread2), hpd_est(spread2)))
colnames(out) <- c("mean", "mode", "sd", "LB", "UB")
rownames(out) <- c("(Intercept)", "sexfemale")

round(out, 2)

```


# Results
\scriptsize
```{r, echo = FALSE}
round(coef_circ(fit, type = "continuous")*(180/pi), 2)
```

# Results

\begin{figure}[H]
        \centering

\includegraphics[width=0.8\textwidth]{plots/FigurelineSBSODreal.png}

		\label{}
\end{figure}

Interpretation of coefficients, see @CremersMulderKlugkist2017. 

# Some thoughts on circular statistics and the Libet clock

```{r, echo = FALSE}
sample_1 <- circular::rvonmises(1000, 2, 10, control.circular = list(units = "hours"))
sample_2 <- circular::rvonmises(1000, 2, 5, control.circular = list(units = "hours"))
sample_3 <- circular::rvonmises(1000, 2, 2, control.circular = list(units = "hours"))
```

```{r, echo = FALSE}
par(mfrow = c(1,3))
plot(sample_1, stack = TRUE, template = "clock12", main = "Group 1")
plot(sample_2, stack = TRUE, template = "clock12", main = "Group 2")
plot(sample_3, stack = TRUE, template = "clock12", main = "Group 3")
```

# Some thoughts on circular statistics and the Libet clock

\centering Location

\vspace{0.5cm}

\begin{columns}[T] % align columns
\begin{column}{.48\textwidth}

\centering Linear analysis:
\vspace{0.5cm}
\begin{itemize}
\item Group 1 mean: `r round(mean(as.numeric(sample_1)), 3)`
\item Group 2 mean: `r round(mean(as.numeric(sample_2)), 3)`
\item Group 3 mean: `r round(mean(as.numeric(sample_3)), 3)`
\end{itemize}

\end{column}%
\hfill%
\begin{column}{.48\textwidth}

\centering Circular analysis:
\vspace{0.5cm}
\begin{itemize}
\item Group 1 mean: `r round(mean(sample_1), 3)`
\item Group 2 mean: `r round(mean(sample_2), 3)`
\item Group 3 mean: `r round(mean(sample_3), 3)`
\end{itemize}

\end{column}%
\end{columns}

# Some thoughts on circular statistics and the Libet clock

\centering Accuracy

\vspace{0.5cm}

\begin{columns}[T] % align columns
\begin{column}{.48\textwidth}

\centering Linear analysis:
\vspace{0.5cm}
\begin{itemize}
\item Scores converted to 0-180 degree scale, deviance from correct time
\item linear regression
\item heterogenous regression models
\end{itemize}

\end{column}%
\hfill%
\begin{column}{.48\textwidth}

\centering Circular analysis:
\vspace{0.5cm}
\begin{itemize}
\item Models that allow for simultaneous modelling of mean and variance (heterogenous)
\item Projected normal regression (available in R-package `bpnreg`)
\item Heterogenous von Mises models
\end{itemize}

\end{column}%
\end{columns}

# Some thoughts on circular statistics and the Libet clock

\begin{columns}[T] % align columns
\begin{column}{.48\textwidth}

Linear analysis:
\vspace{0.5cm}
\begin{itemize}
\item Careful with scale and results
\item Analysis methods easily available
\end{itemize}

\end{column}%
\hfill%
\begin{column}{.48\textwidth}

Circular analysis:
\vspace{0.5cm}
\begin{itemize}
\item Often quite similar to linear analysis
\item Scale is correct
\item Analysis methods not always readily available 
\end{itemize}

\end{column}%
\end{columns}


```{r, echo = FALSE}
# mean(unclass(c(sample_1, sample_2))-pi)
# mean(unclass(c(sample_3, sample_4))-pi)
# 
# mean_circ(c(sample_1, sample_2)-pi)
# mean_circ(c(sample_3, sample_4)-pi)
```


# Software

* R-packages:
  - [`circular`](https://cran.r-project.org/web/packages/circular/index.html): general purpose package
  - [`bpnreg`](https://cran.r-project.org/web/packages/bpnreg/index.html): Bayesian circular regression and mixed-effects models based on projected normal distribution
  - [`circglmbayes`](https://cran.r-project.org/web/packages/circglmbayes/index.html): Bayesian circular regression based on von Mises distribution
  - [`CircSpaceTime`](https://cran.r-project.org/web/packages/CircSpaceTime/index.html): Bayesian spatial and spatio-temporal models.

* Other:
  - [`CircStat`](https://github.com/circstat/circstat-matlab): general purpose MATLAB toolbox


# Suggested literature

* Introductory

  - Batschelet (1981). Circular statistics in biology.
  - Fisher (1995). Statistical analysis of circular data.
  - Pewsey, Neuhäuser \& Ruxton (2013). Circular statistics in R.
  - Ley \& Verdebout (Eds.). (2018). Applied directional statistics: Modern methods and case studies.
  - Cremers \& Klugkist (2018). [One direction? A tutorial for circular data analysis using R with examples in cognitive psychology.](https://www.frontiersin.org/articles/10.3389/fpsyg.2018.02040/full)

* Advanced (requires mathematical background)

  - Pewsey \& García-Portugués (2020). [Recent advances in directional statistics.](https://arxiv.org/abs/2005.06889)
  - Mardia \& Jupp (2000) Directional statistics.
  - Ley \& Verdebout (2017). Modern directional statistics.


# References
